The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-07-30 paper.qmd
added comment to eda
FInished eda section

This t-test, f-test example highlights a fundamental problem with the VSUP and even the bivariate map. When we blend colors to make them visually indistinguishable, we need to decide at what level of *uncertainty* to blend these colors together. There is no intrinsic mapping of "uncertainty", so we need to make a decision as to how much "suppressing" the visualisation does. We did this implicitly in the bivarate map by deciding how much "variance"

- we could consider this a power thing?

Using the methods discussed thus far, where each cell is colours by a single value, it would be impossible to succeed in this method. However, this may not be true if we change our approach entirely.

@@

# Implicitly Combining Uncertainty and Signal
## Pixel maps
```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-pixel
#| fig-cap: "A pixel map of the counties of Iowa. In this map, each county is broken up into several smaller regions and coloured according to a potential daily temperature, given the its average daily temperature and its sampling distribution. This results in each county being represented by a sample rather than a single value."
#| fig-subcap: 
#|   - "High Variance Data"
#|   - "Low Variance Data"
#|   - "Pixel Map Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"

@@

```
## Show me the data

@@

Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticeable to a human at that level. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. These experiments utilised the line-up protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. fig-hypvs shows the conceptual difference between the line-up protocol and uncertainty visualisation.

>>>>

This t-test, f-test example highlights a fundamental problem with the VSUP and even the bivariate map. When we blend colors to make them visually indistinguishable, we need to decide at what level of *uncertainty* to blend these colors together. There is no intrinsic mapping of uncertainty, so we need to make a decision as to how much signal-suppression the visualisation should perform. There is no existing rule or accepted practice in how to go about this, but it is present in every uncertainty graphic. Even though the bivariate map was not explicitly combining temperature values at certain variance levels, the mapping of variance to color saturation does this implicitly. That is, at certain saturation values the colors in a bivariate map are imperceptibly different and appear as though they are mapped to the same value, regardless of whether or not that is actually the case.

Regardless as to whether the map colors are implicitly or explicitly indistinguishable, the point at which the colors blend together can only be validated by a single hypothesis. Therefore, using the methods discussed thus far, it would be impossible to design signal-suppression for EDA. If we are designing a graphic for communication then we can suppress the signal we were seeking to communicate, but this would rely on the distinction between EDA and communication to be somewhat clear cut. Visualization authors may design a graphic with a particular purpose, but what the visualization *is actually* used for is determined by the viewer and some authors have discussed the validity of there being any distinction between visualisation motivation's at all [@Hullman2021]. However, this problem only exists when we color each cell using a single color. If we can color a cell using multiple colors, this limitation may disappear entirely.

@@

# Implicitly Combining Uncertainty and Signal

## Pixel maps

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-pixel
#| fig-cap: "A pixel map of the counties of Iowa. In this map, each county is broken up into several smaller regions and coloured according to a potential daily temperature, given the its average daily temperature and its sampling distribution. This results in each county being represented by a sample rather than a single value."
#| fig-subcap: 
#|   - "High Variance Data"
#|   - "Low Variance Data"
#|   - "Pixel Map Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"

@@

## What is uncertainty?

## Show me the data
@Kim2019 looked at how different uncertainty visualisation techniques influenced the percieved sample size

@@

@Patrick2023 found that human viewers were less sensitive to deviations from the null hypothesis than the typical statistical tests.

What this all means, is that there will always be some trade off when we are designing a visualisation for signal-suppression. There will be a trade off in *which* signal we suppress, as well as *how much* we suppress it. This is true for any graphic that is comparable in some way to a statistical test or process. A good example is the line-up protocol, which is a visualisation tool that used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010]. This motivation is similar to our signal-suppression goal. @Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing. In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference. While these methods apply different statistical philosophies, they both investigated the sensitivity of visualisation methods relative to existing evaluation methods. This approach may prove useful when considering
